Vestibular Contributions to Spatial Perspective Transformations: From Sensory Inference To Imagined Self-Motion

### 2017-01-25 manuscript.Rmd
fixed ref commands and added png versions of diagrams

Consider the use of deictic pronouns, such as "this" or "that" and "left" or "right". These terms only assume a their meaning when placed within a certain frame of reference (FOR). For example, in a dialogue between two speakers, the statement "I would like that one", or "I would like the one on the right" can only be understood under the assumption that the deictic terms "that" and "right" are used in an egocentric FOR. In addition, both interlocutors must be able to access the other's FOR with respect to an external FOR. In other words: In order to give instructions, the speaker must be able to mentally assume the addressee's position in space. A further example that requires a similar spatial transformation is when two people want to perform a coordinated action, such as lifting a heavy object. This requires having knowledge of the other's potential actions in order to adjust one's own actions, and this again requires knowledge of the other's position with an external FOR. A judgement about what kind of actions the other person can perform requires an embodied simulation using a representation of one's own body schema [@Pezzulo:2013iy]. Giving an instruction to another person thus requires running an embodied simulation in order to infer what kind of motor commands the other person should perform __[Blakemore and Decety, 2001; Wolpert et al., 2003; Jeannerod, 2006; Pezzulo et al., 2007, 2013; Dindo et al., 2011; Pezzulo, 2011a,b]__. This embodied simulation is performed using one's own body schema. Communicating this information to the other actor requires knowing the position of the addressee's FOR relative to the external reference frame, and the difference between the speaker and the addressee's egocentric reference frames. Perspective taking thus can be conceived of as a aligning a representation of one's body with that of another person in order to access spatial information from a viewpoint other than one's own egocentric viewpoint. In order to achieve this alignment, one must perform translations and rotations of one's own egocentric reference frame. Processing translations and rotations of the head and body is the core domain of the vestibular system, and thus is of particular interest for research into mental spatial transformations. 
Spatial transformations in 6 dimensions (translation along 3 axes, and rotations around 3 axes) are highly complex, and non-linear, and we propose that the brain adopts the strategy of simulating motion of the body through space in order achieve a desired target position. This effectively amounts to a running a simulation of dynamical system, consisting of simulated limb movements and the resulting sensory consequences. \autoref{fig:spt-schematic} (a) shows a simple scenario in which a speaker wants to give an addressee, standing opposite, verbal instructions about an action to perform on an external object. The speaker must determine the most suitable action, given the addressee's position relative to the external object, and in order to achieve this goal, the speaker must align himself with the addressee, so that he can use his own body-centred reference frame to quickly simulate the best action. In this example, alignment with the addressee's reference frame consists of a forwards translation and a rotation around the vertical axis. This can be described as an imagined movement, in which the goal is achieve a specific position.

@@

```{r spt-schematic, fig.cap="Perspective taking and imagined self-motion involve spatial transformations of a representation of the self. (a) If a speaker's goal is to take the perspective of an addressee, he must first solve the inverse problem of determining the appropriate action, and then run a forwards simulation in order to perform a translation and rotation of his own egocentric reference frame. These operations can be viewed as imagined movements.(b) Focussing on imagined rotation around the vertical axis allows us to simplify the problem, and highlights the fact that perspective taking and mental imagery involve common computations.", echo=FALSE}
knitr::include_graphics('diagrams/fig-1-common-process.tiff')
```

In order to do this, the speaker must first determine what type of movement to perform, and the direction, velocity and duration that lead to the desired outcome. This means that in order to know what to simulate, the speaker must first work backwards from the desired outcome, and infer the actions that are most likely to lead to that outcome. This is known in the field of motor control as the inverse problem [@Wolpert:1998td]. In order to achieve the desired position, the motor commands can then be simulated, and the expected sensory consequences can be estimated; this is known as the forward problem.  \autoref{fig:spt-schematic} (b) shows that if the goal is to simply simulate the perceptual experience of rotation around the body's vertical (yaw) axis, rather than project a representation of one's body onto a target, the problem is simplified; the actor can perform forward inference without the need to determine the optimal motor action. As demonstrated by @Penny:2013iv, these computations can be performed in a common probabilistic model. Interestingly, this has been discussed in the context of reinforcement learning, and planning in humans has been described as reverse inference [@Botvinick:2012fe]. Further, [@Pickering:2014bo] discuss the relationship between control-theoretic notions of inverse/forward models, known collectively as internal models, and and more general concepts of a generative model in a hierarchical Bayesian framework.

@@

## Generative model for sensory inference
We proceed by providing a brief introduction to the concept of state-space models, and particle filtering, which is a particular algorithm that can be used to perform inference in state-space models. While Kalman filters provide the optimal solution when the restrictions of linear Gaussian State-space models are met, they are not easily extensible to non-linear models, which is relevant when considering multi-dimensional motion. Particle filtering can be applied to arbitrary nonlinear models. State-space models can be seen as special cases of more general Bayesian hierarchical models [@Bishop:2006ui; @Murphy:2012ua]. Following the same approach as @Penny:2013iv, we first reframe the generative model used for sensory inference as a dynamic probabilistic graphical model
 <!-- from a statistical perspective, rather an engineering perspective -->
A state-space model is special case of a dynamic latent variable model which is suitable for modelling the time-varying behaviour of a system in terms of latent, unobservable processes and noisy observations. As an example, consider trying to estimate the velocity and position during a leftward head turn. The afferent SCC signals are considered to be proportional to the angular velocity of the head with a frequency band of approximately 0.1 – 10 Hz, which constitutes the range of natural head movements [@Grabherr:2008kk; @Carriot:2014fv]. The measurements are therefore observations of the true, unobservable angular velocity; however, they are contaminted by Gaussian noise, which is determined by characteristics of the sensory organs. \autoref{fig:head-turn-data} shows a sequence of noisy measurements during a two-second turn of the head. The acceleration generating the movement is sinusoidal, given for each time step $t$ by $Asin(2\pi t/T)$, where $A$ is the amplitude of the movement, $T = 1/f$ is the duration of the movement, and $f$ is the frequency. This acceleration results in a peak velocity $\omega_{peak} = AT/\pi$ and a final angular position $\theta_{final} = AT^2/2\pi$.  Due to the inherent noisiness of the measurements, and the dynamic nature of the stimulus, the measurements need to be interpreted according to the context in which they occur.

@@

<!-- \autoref{fig:graphical-model-inference} -->
\autoref{fig:graphical-model-inference} shows a representation of a state-space model as a graphical model, unrolled over multiple time steps. Starting at time 0, the state evolves according to the process model, with the measurements at each time step being generated according to the measurement model. The 2-dimensiontal latent state variable is unshaded, to indicate that it cannot be observed, wheres the observed sensory measurements are shaded. Whereas previous models were discussed in terms of the particular filtering algorithm used, the graphical notation highlights the fact that a state-space model can be viewed as a dynamic latent factor model [@Blei:2014cp]. This opens up the possibility of extending the graphical model to include higher-level variables. 
Given a generative model, we can apply various sampling algorithms, either in order to run a Monte Carlo simulation by sampling from the prior without incorporating any observed variables, or to perform inference, by conditioning on observed variables. After a brief discussion particle filtering, an inference algorithm often used to infer the values of the latent state variables, we will introduce our extension to the model that will allow a prediction of the consequences of active movements.

<!-- ![Graphical model for inferring the 2-dimensional state $x$, consisting of angular velocity and position, based on noisy sensory measurements $y$. ](diagrams/dynamic-belief-network-inference-1.pdf){#fig:graphical-model-inference width=100%} -->

```{r graphical-model-inference, fig.cap="Graphical model for inferring the 2-dimensional state $x$, consisting of angular velocity and position, based on noisy sensory measurements $y$.", echo=FALSE}
knitr::include_graphics('diagrams/dynamic-belief-network-inference-1.tiff')
```

@@

\autoref{fig:particle-filter-explained} shows a visualization of the computations performed by a particle filter at each time step, in order to track the angular velocity. Initially, the particles are propagated forwards in time according to the proposal distribution. They are equally weighted, indicated by their diameters. During the update step, the particle are reweighted, with each particle’s weights reflecting how well it explains the measurement. This is visualized by the change in diameter; those particles lying close to the observation (sensory input from the horizontal canal) have a larger diameter. After resampling, the new set of particles is pulled towards the measurement, and the weights are reset. This adjustment toward the observation depends on both the process noise, and the modelled sensor noise. The uncertainty of the belief, as indicated by the particle cloud's standard deviation (light grey lines), increases at every prediction step, and subsequently decreases after updating. This cycle of prediction, updating and resampling ensures that the collection of particles comes to represent the 'true' filtering distribution, by eliminating those particles with low weights. The particle filter represents all beliefs in the form of probability distributions; the mean and variance of the belief distribution can be computed at each time step, and the logarithm of the weights $log(\frac{1}{N}\sum^N_{i=1}w_i)$, averaged over all $N$ particles, may be summed over all time steps in order to compute the log-likelihood $\sum^T_{t=1}\Big[log(\sum^N_{i=1}w_i) - log(N)\Big]$, which can be used in order to evaluate how well the model was able to fit the sensory evidence. The summed log-likelihood can be used for parameter estimation in extensions of the basic particle filtering algorithm [@Andrieu:2010gc].

@@

```{r particle-filter-explained, fig.cap="Particle filter computations performed at every time step. The cloud of particles (grey circles) represents the current belief state. Initially, a prediction is made for each particle by sampling from the proposal distribution, which is determined by the agent’s kinematic model. This is equivalent to sampling from the prior distribution. Then, during the updating step, an observation  is obtained (shown as a black diamond). Each particle is assigned an importance weight, according to how well it explains the observation. Each particle's diameter reflects its importance weight. Subsequently, a new set of particles is formed by weighted multinomial resampling, with each particle's probability of being chosen proportional to its importance weight. The particle cloud's mean and standard deviation are shown as light grey circles and lines. This resulting cloud of particles represents the posterior distribution. This posterior subsequently forms the prior distribution for the next time step.", echo=FALSE}
knitr::include_graphics('diagrams/figure-particle-filter-explained.tiff')
```

\autoref{fig:passive} shows the result of applying the particle filtering algorithm to the graphical model depicted in \autoref{fig:graphical-model-inference}, in order to infer the velocity and position during a two-second leftward head turn.

@@

This graphical model assumes that the velocity is constant, and does not  the explicitly estimate acceleration; however, it allows for random changes in its velocity by in terms of the process noise. The estimation can be improved by increasing the process noise. \autoref{fig:passive-2} shows the resulting estimates. Whilst the lag is somewhat reduced, a consequence of increasing the process noise is that the measurement are given more weight, resulting in a tendency to overfit the noise in the measurements. The constant velocity (random acceleration) model is unable to capture the underlying smooth trajectory of the velocity and position, instead showing random jumps in the velocity estimate.

@@

In terms of state-space models, this is implemented as a control input $u$, which extends the graphical model shown in \autoref{fig:graphical-model-inference}. In this model, the estimated acceleration is replaced by an known acceleration signal. This capability to predict the consequences of self-initiated movements is essential in order to distinguish re-afference from ex-afference - the system has knowledge about how it moves, and can use this to predict the sensory consequences. A further benefit of including a control input is that the model can now be used in order to run counterfactual simulations of the dynamic behaviour of the system for various types of movements.

@@

\autoref{fig:graphical-model-control} shows the resulting graphical model.

```{r graphical-model-control, fig.cap="Graphical model extended with hierarchical model for self-initiated movements.", echo=FALSE}
knitr::include_graphics('diagrams/dynamic-belief-network-inference-2.tiff')
```

>>>>

Consider the use of deictic pronouns, such as "this" or "that" and "left" or "right". These terms only assume a their meaning when placed within a certain frame of reference (FOR). For example, in a dialogue between two speakers, the statement "I would like that one", or "I would like the one on the right" can only be understood under the assumption that the deictic terms "that" and "right" are used in an egocentric FOR. In addition, both interlocutors must be able to access the other's FOR with respect to an external FOR. In other words: In order to give instructions, the speaker must be able to mentally assume the addressee's position in space. A further example that requires a similar spatial transformation is when two people want to perform a coordinated action, such as lifting a heavy object. This requires having knowledge of the other's potential actions in order to adjust one's own actions, and this again requires knowledge of the other's position with an external FOR. A judgement about what kind of actions the other person can perform requires an embodied simulation using a representation of one's own body schema [@Pezzulo:2013iy]. Giving an instruction to another person thus requires running an embodied simulation in order to infer what kind of motor commands the other person should perform __[Blakemore and Decety, 2001; Wolpert et al., 2003; Jeannerod, 2006; Pezzulo et al., 2007, 2013; Dindo et al., 2011; Pezzulo, 2011a,b]__. This embodied simulation is performed using one's own body schema. Communicating this information to the other actor requires knowing the position of the addressee's FOR relative to the external reference frame, and the difference between the speaker and the addressee's egocentric reference frames. Perspective taking thus can be conceived of as a aligning a representation of one's body with that of another person in order to access spatial information from a viewpoint other than one's own egocentric viewpoint. In order to achieve this alignment, one must perform translations and rotations of one's own egocentric reference frame. Processing translations and rotations of the head and body is the core domain of the vestibular system, and thus is of particular interest for research into mental spatial transformations. 
Spatial transformations in 6 dimensions (translation along 3 axes, and rotations around 3 axes) are highly complex, and non-linear, and we propose that the brain adopts the strategy of simulating motion of the body through space in order achieve a desired target position. This effectively amounts to a running a simulation of dynamical system, consisting of simulated limb movements and the resulting sensory consequences. Figure \@ref(fig:spt-schematic) (a) shows a simple scenario in which a speaker wants to give an addressee, standing opposite, verbal instructions about an action to perform on an external object. The speaker must determine the most suitable action, given the addressee's position relative to the external object, and in order to achieve this goal, the speaker must align himself with the addressee, so that he can use his own body-centred reference frame to quickly simulate the best action. In this example, alignment with the addressee's reference frame consists of a forwards translation and a rotation around the vertical axis. This can be described as an imagined movement, in which the goal is achieve a specific position.

@@

```{r spt-schematic, fig.cap="Perspective taking and imagined self-motion involve spatial transformations of a representation of the self. (a) If a speaker's goal is to take the perspective of an addressee, he must first solve the inverse problem of determining the appropriate action, and then run a forwards simulation in order to perform a translation and rotation of his own egocentric reference frame. These operations can be viewed as imagined movements.(b) Focussing on imagined rotation around the vertical axis allows us to simplify the problem, and highlights the fact that perspective taking and mental imagery involve common computations.", echo=FALSE}
knitr::include_graphics('diagrams/fig-1-common-process.png')
```

In order to do this, the speaker must first determine what type of movement to perform, and the direction, velocity and duration that lead to the desired outcome. This means that in order to know what to simulate, the speaker must first work backwards from the desired outcome, and infer the actions that are most likely to lead to that outcome. This is known in the field of motor control as the inverse problem [@Wolpert:1998td]. In order to achieve the desired position, the motor commands can then be simulated, and the expected sensory consequences can be estimated; this is known as the forward problem.  Figure \@ref(fig:spt-schematic) (b) shows that if the goal is to simply simulate the perceptual experience of rotation around the body's vertical (yaw) axis, rather than project a representation of one's body onto a target, the problem is simplified; the actor can perform forward inference without the need to determine the optimal motor action. As demonstrated by @Penny:2013iv, these computations can be performed in a common probabilistic model. Interestingly, this has been discussed in the context of reinforcement learning, and planning in humans has been described as reverse inference [@Botvinick:2012fe]. Further, [@Pickering:2014bo] discuss the relationship between control-theoretic notions of inverse/forward models, known collectively as internal models, and and more general concepts of a generative model in a hierarchical Bayesian framework.

@@

## Generative model for sensory inference
We proceed by providing a brief introduction to the concept of state-space models, and particle filtering, which is a particular algorithm that can be used to perform inference in state-space models. While Kalman filters provide the optimal solution when the restrictions of linear Gaussian State-space models are met, they are not easily extensible to non-linear models, which is relevant when considering multi-dimensional motion. Particle filtering can be applied to arbitrary nonlinear models. State-space models can be seen as special cases of more general Bayesian hierarchical models [@Bishop:2006ui; @Murphy:2012ua]. Following the same approach as @Penny:2013iv, we first reframe the generative model used for sensory inference as a dynamic probabilistic graphical model
 <!-- from a statistical perspective, rather an engineering perspective -->
A state-space model is special case of a dynamic latent variable model which is suitable for modelling the time-varying behaviour of a system in terms of latent, unobservable processes and noisy observations. As an example, consider trying to estimate the velocity and position during a leftward head turn. The afferent SCC signals are considered to be proportional to the angular velocity of the head with a frequency band of approximately 0.1 – 10 Hz, which constitutes the range of natural head movements [@Grabherr:2008kk; @Carriot:2014fv]. The measurements are therefore observations of the true, unobservable angular velocity; however, they are contaminted by Gaussian noise, which is determined by characteristics of the sensory organs. Figure \@ref(fig:head-turn-data) shows a sequence of noisy measurements during a two-second turn of the head. The acceleration generating the movement is sinusoidal, given for each time step $t$ by $Asin(2\pi t/T)$, where $A$ is the amplitude of the movement, $T = 1/f$ is the duration of the movement, and $f$ is the frequency. This acceleration results in a peak velocity $\omega_{peak} = AT/\pi$ and a final angular position $\theta_{final} = AT^2/2\pi$.  Due to the inherent noisiness of the measurements, and the dynamic nature of the stimulus, the measurements need to be interpreted according to the context in which they occur.

@@

<!-- \autoref{fig:graphical-model-inference} -->
\@ref(fig:graphical-model-inference) shows a representation of a state-space model as a graphical model, unrolled over multiple time steps. Starting at time 0, the state evolves according to the process model, with the measurements at each time step being generated according to the measurement model. The 2-dimensiontal latent state variable is unshaded, to indicate that it cannot be observed, wheres the observed sensory measurements are shaded. Whereas previous models were discussed in terms of the particular filtering algorithm used, the graphical notation highlights the fact that a state-space model can be viewed as a dynamic latent factor model [@Blei:2014cp]. This opens up the possibility of extending the graphical model to include higher-level variables. 
Given a generative model, we can apply various sampling algorithms, either in order to run a Monte Carlo simulation by sampling from the prior without incorporating any observed variables, or to perform inference, by conditioning on observed variables. After a brief discussion particle filtering, an inference algorithm often used to infer the values of the latent state variables, we will introduce our extension to the model that will allow a prediction of the consequences of active movements.

<!-- ![Graphical model for inferring the 2-dimensional state $x$, consisting of angular velocity and position, based on noisy sensory measurements $y$. ](diagrams/dynamic-belief-network-inference-1.png){#fig:graphical-model-inference width=100%} -->

```{r graphical-model-inference, fig.cap="Graphical model for inferring the 2-dimensional state $x$, consisting of angular velocity and position, based on noisy sensory measurements $y$.", echo=FALSE}
knitr::include_graphics('diagrams/dynamic-belief-network-inference-1.png')
```

@@

Figure \@ref(fig:particle-filter-explained) shows a visualization of the computations performed by a particle filter at each time step, in order to track the angular velocity. Initially, the particles are propagated forwards in time according to the proposal distribution. They are equally weighted, indicated by their diameters. During the update step, the particle are reweighted, with each particle’s weights reflecting how well it explains the measurement. This is visualized by the change in diameter; those particles lying close to the observation (sensory input from the horizontal canal) have a larger diameter. After resampling, the new set of particles is pulled towards the measurement, and the weights are reset. This adjustment toward the observation depends on both the process noise, and the modelled sensor noise. The uncertainty of the belief, as indicated by the particle cloud's standard deviation (light grey lines), increases at every prediction step, and subsequently decreases after updating. This cycle of prediction, updating and resampling ensures that the collection of particles comes to represent the 'true' filtering distribution, by eliminating those particles with low weights. The particle filter represents all beliefs in the form of probability distributions; the mean and variance of the belief distribution can be computed at each time step, and the logarithm of the weights $log(\frac{1}{N}\sum^N_{i=1}w_i)$, averaged over all $N$ particles, may be summed over all time steps in order to compute the log-likelihood $\sum^T_{t=1}\Big[log(\sum^N_{i=1}w_i) - log(N)\Big]$, which can be used in order to evaluate how well the model was able to fit the sensory evidence. The summed log-likelihood can be used for parameter estimation in extensions of the basic particle filtering algorithm [@Andrieu:2010gc].

@@

```{r particle-filter-explained, fig.cap="Particle filter computations performed at every time step. The cloud of particles (grey circles) represents the current belief state. Initially, a prediction is made for each particle by sampling from the proposal distribution, which is determined by the agent’s kinematic model. This is equivalent to sampling from the prior distribution. Then, during the updating step, an observation  is obtained (shown as a black diamond). Each particle is assigned an importance weight, according to how well it explains the observation. Each particle's diameter reflects its importance weight. Subsequently, a new set of particles is formed by weighted multinomial resampling, with each particle's probability of being chosen proportional to its importance weight. The particle cloud's mean and standard deviation are shown as light grey circles and lines. This resulting cloud of particles represents the posterior distribution. This posterior subsequently forms the prior distribution for the next time step.", echo=FALSE}
knitr::include_graphics('diagrams/figure-particle-filter-explained.png')
```

Figure \@ref(fig:passive) shows the result of applying the particle filtering algorithm to the graphical model depicted in Figure \@ref(fig:graphical-model-inference), in order to infer the velocity and position during a two-second leftward head turn.

@@

This graphical model assumes that the velocity is constant, and does not  the explicitly estimate acceleration; however, it allows for random changes in its velocity by in terms of the process noise. The estimation can be improved by increasing the process noise. Figure \@ref(fig:passive-2) shows the resulting estimates. Whilst the lag is somewhat reduced, a consequence of increasing the process noise is that the measurement are given more weight, resulting in a tendency to overfit the noise in the measurements. The constant velocity (random acceleration) model is unable to capture the underlying smooth trajectory of the velocity and position, instead showing random jumps in the velocity estimate.

@@

In terms of state-space models, this is implemented as a control input $u$, which extends the graphical model shown in Figure \@ref(fig:graphical-model-inference). In this model, the estimated acceleration is replaced by an known acceleration signal. This capability to predict the consequences of self-initiated movements is essential in order to distinguish re-afference from ex-afference - the system has knowledge about how it moves, and can use this to predict the sensory consequences. A further benefit of including a control input is that the model can now be used in order to run counterfactual simulations of the dynamic behaviour of the system for various types of movements.

@@

Figure \@ref(fig:graphical-model-control) shows the resulting graphical model.

```{r graphical-model-control, fig.cap="Graphical model extended with hierarchical model for self-initiated movements.", echo=FALSE}
knitr::include_graphics('diagrams/dynamic-belief-network-inference-2.png')
```
